---
title: Motor Trend analysis - is manual transmission or automatic transmission better
  for mpg?
author: "Filip Wójcik"
date: "Monday, May 18, 2015"
output: 
  pdf_document:
    keep_tex: true
fontsize: 10pt
---

```{r echo=FALSE, include=FALSE}
library(ggplot2)
library(GGally)
library(reshape)
data(mtcars)
```

# Summary

After performing analysis, the cocnclusion is, that manual transmission results in a bigger mpg value and consequently is a better option. Change from automatic to manual transmission results in a approximately 2.9358 miles/gallon increase (when taken into account along with vehicle weight and calculate on the 1/4 mile time).

# Exploratory data analysis

Exploratory data analysis plots (for each variable) can be found as **Figure1** in an appendix.

# Is an automatic or manual transmission better for MPG

**Figure2 (boxplot of mpg means between two groups)** and **figure3 (correlation plot)** from appendix shows the difference in mpg values between two groups - with manual and automatic transmission. From the plot we can conclude that **variances are not equal between the groups**. We can **suppose** (judging the data on the plot by eye) that the manual transmission system comes with higher values of mpg.  This leads to two hypotheses:

H0: **mean mpg value is equal** for manual and automatic transmission system

Ha: **mean mpg value is higher** for the manual transmission system

We can perform the following t-test for those:

```{r t_test}
auto.mtcars <- mtcars[mtcars$am == 0, "mpg"]
manual.mtcars <- mtcars[mtcars$am == 1, "mpg"]
t.test(manual.mtcars, auto.mtcars, paired=FALSE, var.equal=FALSE, alternative = "greater")
```

With p-value of 0.0006868 we can safely reject the null hypothesis and assume, that with manual transmission mpg is higher.

Because will all would like to travel more with the same amout of fuel, we can conclude, that **manual transmission** is better for mpg.

# Quantify the MPG difference between automatic and manual transmissions

To fill this task, regression analysis will be made. The **slope coefficient for "am"** (transmission type) will describe the difference between two types of transmissions. 
Let's try a simplest possible model:
```{r simple_model}
summary(lm(data=mtcars, formula=mpg ~ am))
```
P-value of **0.000285** suggests that **am** has significant impact on **mpg** but, this model explains only 35.9% of variance (R-squared, used because with am only model is a single linear regression). More variables (x values) will need to be used, to fill this gap.

Because there are multiple variables, that can (potentially) affect mpg value, a proper model needs to be selected from all possible options. Hovewer, some variables affecting mpg may be **multiplecolinear** - so related one with each other. We have to eliminate them, because they are redundant.

Space limitations for this report does not allow to perform the full analysis step-by-step, but the **algorithm of backwards elimination ** proceeds as follows:

1. Find multiple regression coefficients for the full model (all variables)
2. Remove each single variable and re-generate models, quantify residuals and adjusted R^2
3. Generate all possible combinations of 2 variables, iterate through them, and regenerate model, quantify residuals, adjusted R^2 and Akiko Information Criterion (AIC)
... (proceed with combinations of 3 variables, etc. until only single variable remains)
4. From all the models - select the one with the best AIC (best-fitting model)

```{r stepwise_model_selection, results='hide'}
step(object = lm(mtcars$mpg ~ ., data= mtcars), direction = "backward")
```

The final model selected by stepwise, backwards elimination algorithm is:

**mtcars$mpg ~ wt + qsec + am**

```{r multilinear_model_summary, echo=FALSE}
model <- lm(data=mtcars, formula = mpg ~ wt + qsec + am)
summary(model)
```

This model explains 83.3% of variance and indicates that

**Changing from automatic to manual transmission mode changes the mpg by 2.9358**

## Appendix

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='left'}
d <- melt(mtcars)
ggplot(d,aes(x = value)) + 
    facet_wrap(~variable,scales = "free_x") + 
    geom_histogram() + ggtitle("Figure 1 - exploratory data analysis plot")
```

```{r mpg_automatic_vs_manual, echo=FALSE, fig.align='left'}
ggplot(mtcars, aes(x=factor(am), y=mpg)) + geom_boxplot() + ggtitle("Figure 2 - mpg values in groups with manual and automatic transmission")
```

```{r mpg_am_correlation_plot, fig.align='left', echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = mtcars, aes(x=am, y=mpg), ) + geom_line() + ggtitle("Figure 3 - mpg and am correlation plot")
```

```{r final_model_plot_correlation, echo=FALSE, message=FALSE, warning=FALSE}
GGally::ggpairs(data = mtcars[,c("mpg", "wt", "qsec", "am")], diag=list(continuous="bar", params=c(colour="blue")), lower=list(continuous="smooth"), title="Figure 4-final model plot. Variables correlations")
```

```{r final_model_plot_fit_results, echo=FALSE, message=FALSE, warning=FALSE, fig.align='left'}
qplot(x = model$fitted.values, y=model$residuals, xlab = "fitted values", ylab="residuals") + geom_point() + geom_smooth(method="lm") + ggtitle("Figure 5- final model residuals plot")
```